---
title: "Stat531_final"
output: html_document
date: "2024-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
data = read_csv("TB_data_usa .csv",show_col_types = FALSE)
case_rate = data$Number...2
```

```{r load data, echo=FALSE}
#data = read_csv("TB_data_usa .csv",show_col_types = FALSE)
#case_rate = data$Number...2
```

### Explore periods {.unnumbered}

```{r decompose, echo=FALSE}
par(mar=c(6, 4, 4, 2))
pc_spectrum_smooth <- spectrum(case_rate, spans = c(5, 7, 5), 
                       main = "Smoothed periodogram")

mtext(text = expression(bold("Figure:") ~ "incidence rate for tuberculosis cases with max density frequency point."), side = 1, line = 5,cex = 0.75,  adj = 0.5)

max_density_freq = round(pc_spectrum_smooth$freq[which.max(pc_spectrum_smooth$spec)], 3)

abline(v = max_density_freq, lty = "dashed", col = "red", lwd = 2)

text(x = 0.09, y = 5e+1, labels = sprintf("Max: %.3f", max_density_freq), col = "red")

```

The plot shows no periodicity for our data, so SARIMA model should be excluded for analysis. Instead, ARIMA model should be considered  
given that our data clearly have a decreasing trend.



```{r select_arima__module, echo=FALSE,results='hide',warning=FALSE}
model_selection_table <- function(data, 
                                  max_p, d, max_q, 
                                  P = 0, D = 0, Q = 0, period = 0,
                                  simulation_times = 100) {
  aic_table <- matrix(NA, max_p + 1, max_q + 1)
  smallest_root_table <- matrix(NA, max_p + 1, max_q + 1)
  fisher_ci_cover_0_table <- matrix(NA, max_p + 1, max_q + 1)
  simulated_ci_cover_0_table <- matrix(NA, max_p + 1, max_q + 1)
  residual_normal_test_table <- matrix(NA, max_p + 1, max_q + 1)
  residual_acf_outlier_table <- matrix(NA, max_p + 1, max_q + 1)
  
  dimnames(aic_table) <- list(paste("AR", 0:max_p, sep = ""),
                            paste("MA", 0:max_q, sep = ""))
  dimnames(smallest_root_table) <- list(paste("AR", 0:max_p, sep = ""),
                            paste("MA", 0:max_q, sep = ""))
  dimnames(fisher_ci_cover_0_table) <- list(paste("AR", 0:max_p, sep = ""),
                            paste("MA", 0:max_q, sep = ""))
  dimnames(simulated_ci_cover_0_table) <- list(paste("AR", 0:max_p, sep = ""),
                            paste("MA", 0:max_q, sep = ""))
  dimnames(residual_normal_test_table) <- list(paste("AR", 0:max_p, sep = ""),
                            paste("MA", 0:max_q, sep = ""))
  dimnames(residual_acf_outlier_table) <- list(paste("AR", 0:max_p, sep = ""),
                            paste("MA", 0:max_q, sep = ""))
  
  is_sarima_model <- any(c(P, D, Q) != 0)
  has_intercept <- (d == 0) && (D == 0)
  for (p in 0:max_p) {
    for (q in 0:max_q) {
      has_at_least_one_param <- ((p + q + P + Q) > 0)
       
      if (is_sarima_model) {
        pc_model <- try(arima(data, order = c(p, d, q),
                           seasonal = list(order = c(P, D, Q), period = period)),
                        silent = TRUE)
      } else {
        pc_model <- try(arima(data, order = c(p, d, q)),
                        silent = TRUE)
      }
      
      if (inherits(pc_model, "try-error")) {
          next
      }
      
      # aic
      aic_table[p + 1, q + 1] <- pc_model$aic
      
      # smallest root
      if (p != 0) {
        pc_model_ar_root <- polyroot(c(1, -coef(pc_model)[paste0("ar", 1:p)]))
        pc_model_ar_root_mod <- Mod(pc_model_ar_root)
      } else {
        pc_model_ar_root <- NA
        pc_model_ar_root_mod <- NA
      }
      
      if (q != 0) {
        pc_model_ma_root <- polyroot(c(1, coef(pc_model)[paste0("ma", 1:q)]))
        pc_model_ma_root_mod <- Mod(pc_model_ma_root)
      } else {
        pc_model_ma_root <- NA
        pc_model_ma_root_mod <- NA
      }
      
      if (P != 0) {
        pc_model_sar_root <- polyroot(c(1, -coef(pc_model)[paste0("sar", 1:P)]))
        pc_model_sar_root_mod <- Mod(pc_model_sar_root)
      } else {
        pc_model_sar_root <- NA
        pc_model_sar_root_mod <- NA
      }
      
      if (Q != 0) {
        pc_model_sma_root <- polyroot(c(1, coef(pc_model)[paste0("sma", 1:Q)]))
        pc_model_sma_root_mod <- Mod(pc_model_sma_root)
      } else {
        pc_model_sma_root <- NA
        pc_model_sma_root_mod <- NA
      }
      
      if (any(c(p, q, P, Q) != 0)) {
        smallest_root_table[p + 1, q + 1] <- min(c(pc_model_ar_root_mod,
                                                 pc_model_ma_root_mod,
                                                 pc_model_sar_root_mod,
                                                 pc_model_sma_root_mod), na.rm = TRUE)
      }
      
      # fisher ci
      if (any(c(p, q, P, Q) != 0)) {
        fisher_ci_low <- pc_model$coef - 1.96 * diag(pc_model$var.coef)
        fisher_ci_high <- pc_model$coef + 1.96 * diag(pc_model$var.coef)
        
        if (has_intercept) {
          fisher_ci_low <- fisher_ci_low[1:(length(fisher_ci_low) - 1)]
          fisher_ci_high <- fisher_ci_high[1:(length(fisher_ci_high) - 1)]
        }
        
        if (any(fisher_ci_low <= 0 & fisher_ci_high >= 0)) {
          fisher_ci_cover_0_table[p + 1, q + 1] <- TRUE
        } else {
          fisher_ci_cover_0_table[p + 1, q + 1] <- FALSE
        }
      }
      
      # simulated ci
      if (has_at_least_one_param && (simulation_times > 0)) {
        if (!is_sarima_model) {
          simulated_ci <- simulation_arima(pc_model, c(p, d, q), 
                                           has_intercept, simulation_times, length(data))
        } else {
          simulated_ci <- simulation_sarima(pc_model, c(p, d, q), c(P, D, Q), period,
                                            has_intercept, simulation_times, length(data))
        }
        
        if (!any(is.na(simulated_ci))) {
          simulated_ci_low <- simulated_ci[1, ]
          simulated_ci_high <- simulated_ci[2, ]
          if (has_intercept) {
            simulated_ci_low <- simulated_ci_low[1:(length(simulated_ci_low) - 1)]
            simulated_ci_high <- simulated_ci_high[1:(length(simulated_ci_high) - 1)]
          }
          
          if (any(simulated_ci_low <= 0 & simulated_ci_high >= 0)) {
            simulated_ci_cover_0_table[p + 1, q + 1] <- TRUE
          } else {
            simulated_ci_cover_0_table[p + 1, q + 1] <- FALSE
          }
        }
      }
     
      # residual normal test
      shapiro_test_result <- shapiro.test(pc_model$residuals)
      if (shapiro_test_result$p.value < 0.05) {
        residual_normal_test_table[p + 1, q + 1] <- FALSE
      } else {
        residual_normal_test_table[p + 1, q + 1] <- TRUE
      }
      
      # residual acf test
      residual_acf <- acf(pc_model$residuals, plot = FALSE, lag.max = 30)
      acf_ci_high <- qnorm((1 + 0.95) / 2) / sqrt(residual_acf$n.used)
      acf_ci_low <- -acf_ci_high
      residual_acf_outlier_table[p + 1, q + 1] <- sum((residual_acf$acf < acf_ci_low) |
                                                  (residual_acf$acf > acf_ci_high)) - 1
    }
  }
  
  result_table_list <- list(aic_table = aic_table, 
                            smllest_root_table = smallest_root_table,
                            fisher_ci_cover_0_table = fisher_ci_cover_0_table, 
                            simulated_ci_cover_0_table = simulated_ci_cover_0_table,
                            residual_normal_test_table = residual_normal_test_table, 
                            residual_acf_outlier_table = residual_acf_outlier_table)
  
  return(result_table_list)
}


```

```{r, graphsecho=FALSE,warning=FALSE}
select_arima_case <- model_selection_table(case_rate,
                                           max_p = 5, d = 1, max_q = 5,
                                           P = 0, D = 0, Q = 0, period = 0,
                                           simulation_times = 0)
```

```{r, echo=FALSE,results='hide',warning=FALSE}
build_and_diagnose_model <- function(data, model_name, 
                                     arima_order, 
                                     seasonal = FALSE, 
                                     seasonal_order = c(0, 0, 0), period = NULL,
                                     xreg = NULL,
                                     without_summary = FALSE,
                                     without_plot = FALSE) {
  if (seasonal) {
    pc_model <- arima(data, order = arima_order, 
                           seasonal = list(order = seasonal_order, period = period),
                      xreg = xreg)
  } else {
    pc_model <- arima(data, order = arima_order, xreg = xreg)
  }
  
  if (!without_summary) {
    print(pc_model)
  }
  
  p <- arima_order[1]
  d <- arima_order[2]
  q <- arima_order[3]
  P <- seasonal_order[1]
  D <- seasonal_order[2]
  Q <- seasonal_order[3]
  
  if (p > 0) {
    pc_model_ar_roots <- polyroot(c(1, -coef(pc_model)[paste0("ar", 1:p)]))
    cat("AR roots:", round(pc_model_ar_roots, 4), "\n")
    cat("Mod of AR roots:", round(Mod(pc_model_ar_roots), 4), "\n")
  }
  
  if (q > 0) {
    pc_model_ma_roots <- polyroot(c(1, coef(pc_model)[paste0("ma", 1:q)]))
    cat("MA roots:", round(pc_model_ma_roots, 4), "\n")
    cat("Mod of MA roots:", round(Mod(pc_model_ma_roots), 4), "\n")
  }
  
  if (P > 0) {
    pc_model_sar_roots <- polyroot(c(1, -coef(pc_model)[paste0("sar", 1:P)]))
    cat("SAR roots:", round(pc_model_sar_roots, 4), "\n")
    cat("Mod of SAR roots:", round(Mod(pc_model_sar_roots), 4), "\n")
  }
  
  if (Q > 0) {
    pc_model_sma_roots <- polyroot(c(1, coef(pc_model)[paste0("sma", 1:Q)]))
    cat("SMA roots:", round(pc_model_sma_roots, 4), "\n")
    cat("Mod of SMA roots:", round(Mod(pc_model_sma_roots), 4), "\n")
  }
  
  if (!without_plot) {
     par(mfrow = c(2, 2))
     plot(pc_model$residuals, ylab = "Residuals", main = "Residual Plot")
     qqnorm(pc_model$residuals, ylab = "Residuals", main = "QQ Plot")
     qqline(pc_model$residuals)
     acf(pc_model$residuals, type = "correlation",
          lag.max = 40, main = "Autocorrelation Function")
  }
 
  return(invisible(pc_model))
}
```

### ARIMA model building and model diagnostics {.unnumbered}

based on the AIC and smallest root, the relatively suitable model we choose is ARIMA(0,1,5) with smallest root as 1.05
```{r, graphsecho=FALSE,warning=FALSE}
library(knitr)
kable(select_arima_case$aic_table, digits = 3, caption = "AIC of some ARIMA models (incidence rate)")
kable(select_arima_case$smllest_root_table, digits = 4, caption = "Smallest roots of ARIMA models (incidence rate)")
```
To determine invertibility, we need to ensure that all MA roots lie outside the unit circle. Because all the modulus of MA roots are larger than 1, we can say that the moving average part is also invertible. As a result, our ARIMA(0,1,5) can be reasonable treated as an invertible model. 

As for its residuals, we find the residuals seems to be white noise. One thing worth noting is that residuals at earlier time have larger variance than residuals at later time. Therefore, homoscedasticity is violated. Heteroscedasticity does not bias the coefficient estimates themselves, but it can lead to biased estimates of the standard errors, which in turn can mislead inferences made regarding the significance of predictors. In the QQ plot, a heavy-tailed distribution is obvious and this shows that the residual doesn’t obey normal distribution. In the ACF plot, nothing significant is observed, therefore, we can conclude that the residual is uncorrelated. Overall, the diagnosis conducted here suggests that a simple ARIMA model is not a very suitable model for tuberculosis incidence data because the violation of normality and homoscedasticity makes the statistical inference hard to proceed.

```{r fig.width=10, fig.height=5, echo=FALSE}

build_and_diagnose_model(case_rate, 
                         model_name = "ARIMA(2, 1, 0) (incidence rate)",
                         arima_order = c(0, 1, 5),
                         without_summary = TRUE)
```